A fully labelled proof system for intuitionistic modal logics

### 2019-10-03 caleidoscope/system.tex

\documentclass[a4paper]{article}

%\usepackage{aiml18macro}

\usepackage[margin=2cm]{geometry}
\usepackage[noxy]{virginialake}                        
\usepackage{lmodern}
\usepackage{amssymb,amsmath,amsthm} 
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{pgf}
\usepackage{array}
\usepackage[all]{xy}
\usepackage{tikz}
%\usepackage{hyperref}
\usepackage{pgffor}

	%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lem}[theorem]{Lemma}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
	%%
\theoremstyle{definition}
\newtheorem{observation}[theorem]{Observation}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{construction}[theorem]{Construction}
\newtheorem{remark}[theorem]{Remark}
	%%

\tikzset{
	annotated cuboid/.pic={
		\tikzset{%
			every edge quotes/.append style={midway, auto},
			/cuboid/.cd,
			#1
		}
		\draw [every edge/.append style={pic actions, opacity=.5}, pic actions]
		(0,0,0) coordinate (o) -- ++(-\cubescale*\cubex,0,0) coordinate (a) -- ++(0,-\cubescale*\cubey,0) coordinate (b) edge coordinate [pos=1] (g) ++(0,0,-\cubescale*\cubez)  -- ++(\cubescale*\cubex,0,0) coordinate (c) -- cycle
		(o) -- ++(0,0,-\cubescale*\cubez) coordinate (d) -- ++(0,-\cubescale*\cubey,0) coordinate (e) edge (g) -- (c) -- cycle
		(o) -- (a) -- ++(0,0,-\cubescale*\cubez) coordinate (f) edge (g) -- (d) -- cycle;

		;
	},
	/cuboid/.search also={/tikz},
	/cuboid/.cd,
	width/.store in=\cubex,
	height/.store in=\cubey,
	depth/.store in=\cubez,
	units/.store in=\cubeunits,
	scale/.store in=\cubescale,
	width=10,
	height=10,
	depth=10,
	units=cm,
	scale=.1,
}
%tikz parameters

\tikzstyle{point}=[circle,draw]
\usetikzlibrary{arrows,automata,shapes,decorations.markings,
	decorations.pathmorphing,backgrounds,fit,snakes,calc}

%\usepackage{float}
%\floatstyle{boxed} 
%\restylefloat{figure}

\usepackage{colonequals}
\usepackage{bm}

%%% Extracting symbols from MnSymbol 
\DeclareFontFamily{U} {MnSymbolC}{}
%
\DeclareFontShape{U}{MnSymbolC}{m}{n}{
	<-6>  MnSymbolC5
	<6-7>  MnSymbolC6
	<7-8>  MnSymbolC7
	<8-9>  MnSymbolC8
	<9-10> MnSymbolC9
	<10-12> MnSymbolC10
	<12->   MnSymbolC12}{}
\DeclareFontShape{U}{MnSymbolC}{b}{n}{
	<-6>  MnSymbolC-Bold5
	<6-7>  MnSymbolC-Bold6
	<7-8>  MnSymbolC-Bold7
	<8-9>  MnSymbolC-Bold8
	<9-10> MnSymbolC-Bold9
	<10-12> MnSymbolC-Bold10
	<12->   MnSymbolC-Bold12}{}
%
\DeclareSymbolFont{MnSyC}         {U}  {MnSymbolC}{m}{n}
%
\DeclareMathSymbol{\diamondplus}{\mathbin}{MnSyC}{124}
\DeclareMathSymbol{\boxtimes}{\mathbin}{MnSyC}{117}

%%% Comments  
\definecolor{notgreen}{rgb}{.1,.6,.1}
\newcommand{\marianela}[1]{{\color{purple}[Marianela: #1]}}
\newcommand{\sonia}[1]{{\color{blue}[Sonia: #1]}}
\newcommand{\lutz}[1]{{\color{notgreen}[Lutz: #1]}}
\newcommand{\todo}[1]{{\color{red}[TODO: #1]}}

%%% General
\input{macros}
\newcommand{\B}{\mathcal{R}}
\newcommand*{\A}{\mathcal{A}}

\newcommand{\lseq}[3]{#1 , #2 \SEQ #3}

%%% Systems
\newcommand*{\ax}[1]{\mathsf{#1}}
\newcommand*{\kax}[1][]		{\ax{k_{#1}}}
\newcommand*{\lab}{\mathsf{lab}}
\newcommand*{\IK}{\mathsf{IK}}
\newcommand*{\K}{\mathsf{K}}
\newcommand*{\labIKp}{\lab\IK_{\le}}
\newcommand*{\ISfour}{\mathsf{IS4}}
\newcommand*{\Sfour}{\mathsf{S4}}

\newcommand{\fourax}{\mathsf{4}}
%%% Connectives
\newcommand*{\NOT}{\neg}
\newcommand*{\AND}{\mathbin{\wedge}}
\newcommand*{\TOP}{\mathord{\top}}
\newcommand*{\OR}{\mathbin{\vee}}
\newcommand*{\BOT}{\mathord{\bot}}
\newcommand*{\IMP}{\mathbin{\scalebox{.9}{\raise.2ex\hbox{$\supset$}}}}

\newcommand*{\BOX}{\mathord{\Box}}
\newcommand*{\DIA}{\mathord{\Diamond}}

%%% Labelled sequents
\newcommand*{\fm}[1]{{\color{notgreen}{#1}}}
\newcommand*{\lb}[1]{{\color{blue}{#1}}}
\newcommand*{\labels}[2]{\lb{#1}\mathord{:}\fm{#2}}
\newcommand*{\accs}[2]{\lb{#1}R\lb{#2}}
\newcommand*{\futs}[2]{\lb{#1}\le{\color{blue}{#2}}}
\newcommand{\SEQ}{\Longrightarrow}
\newcommand*{\DD}{\mathcal{D}}
\newcommand*{\rn}[1]  {\ensuremath{\mathsf{#1}}}
\newcommand*{\invr}[1]{#1^\bullet}
\newcommand*{\rel}{R}

\newcommand*{\SG}{\fm{\mathcal{G}}}
\newcommand{\SGi}[1]{\fm{\mathcal{G}_{#1}}}
\newcommand{\proviso}[1]{\mbox{\scriptsize #1}}
%%% Labelled rules
\newcommand*{\labrn}[2][]  {\rn{#2}_{#1}}%^{\lab}}}
\newcommand*{\rlabrn}[2][]  {\rn{#2}_\rn{R#1}}%^\lab}}
\newcommand*{\llabrn}[2][]  {\rn{#2}_\rn{L#1}}%^\lab}}
%
\newcommand*{\brsym}{\boxtimes}%\mathord{\scalebox{.8}{$\blacksquare$}}}
\newcommand*{\diasym}{\diamondplus}%\mathord{\blacklozenge}}
%
\newcommand*{\boxbrn}[1]{\brsym_\rn{#1}}%^{\lab}}}
\newcommand*{\diabrn}[1]{\diasym_\rn{#1}}

\newcommand{\inter}[1]{\lb{\llbracket #1\rrbracket}}
\newcommand{\force}[3]{#1,#2\Vdash#3}
\newcommand{\nforce}[3]{#1,#2\not\Vdash#3}
\newcommand{\cforce}[3]{#1,\lb{#2}\Vdash\fm{#3}}
\newcommand{\cnforce}[3]{#1,\lb{#2}\not\Vdash\fm{#3}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Equation environments
\newdimen\mydisplayskip
\mydisplayskip=.4\abovedisplayskip
\newenvironment{smallequation}
{\par\nobreak\vskip\mydisplayskip\noindent\bgroup\small\csname equation\endcsname}{\csname endequation\endcsname\egroup}
\newenvironment{smallequation*}
{\par\nobreak\vskip\mydisplayskip\noindent\bgroup\small\csname equation*\endcsname}{\csname endequation*\endcsname\egroup}
\newenvironment{smallalign}
{\par\nobreak\noindent\bgroup\small\csname align\endcsname}{\csname endalign\endcsname\egroup}
\newenvironment{smallalign*}
{\par\nobreak\noindent\bgroup\small\csname align*\endcsname}{\csname endalign*\endcsname\egroup}
%
\newcommand{\vlderivationauxnc}[1]{#1{\box\derboxone}\vlderivationterm}
\newcommand{\vlderivationnc}{\vlderivationinit\vlderivationauxnc}
%
%
\makeatletter
\newbox\@conclbox
\newdimen\@conclheight
%
%%%%%%
\newcommand{\vlhtr}[2]{\vlpd{#1}{}{#2}}
\newcommand\vlderiibase[5]{{%
		\setbox\@conclbox=\hbox{$#3$}\relax%
		\@conclheight=\ht\@conclbox%
		\setbox\@conclbox=\hbox{$%
			\vlderivationnc{%
				\vliin{#1}{#2}{\box\@conclbox}{#4}{#5}%
			}$}%
		\lower\@conclheight\box\@conclbox%
	}}
%%
\newcommand\vlderibase[4]{{%
		\setbox\@conclbox=\hbox{$#3$}\relax%
		\@conclheight=\ht\@conclbox%
		\setbox\@conclbox=\hbox{$%
			\vlderivationnc{%
				\vlin{#1}{#2}{\box\@conclbox}{#4}%
			}$}%
		\lower\@conclheight\box\@conclbox%
	}}
%
\newcommand\vlderidbase[4]{{%
		\setbox\@conclbox=\hbox{$#3$}\relax%
		\@conclheight=\ht\@conclbox%
		\setbox\@conclbox=\hbox{$%
			\vlderivationnc{%
				\vlid{#1}{#2}{\box\@conclbox}{#4}%
			}$}%
		\lower\@conclheight\box\@conclbox%
	}}
	%
	\makeatother
	%
%%%General maths
\newcommand*{\reducesto}{\quad{\leadsto}\quad}

%%%%%%%%%%%%%%%%%

%\newcommand{\height}[1]{\mathrm{ht}(#1)}
\newcommand{\height}[1]{|#1|}

\newcommand{\quand}{\quad\mbox{and}\quad}
\newcommand{\qquand}{\qquad\mbox{and}\qquad}
\newcommand{\quadcm}{\rlap{\quad,}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%The following line defines the page header consisting of the surnames of the authors.
% Please include only the last names! 
% Separate by commas except the last two surnames which are separated by an "and".
\def\lastname{Marin, Morales and Stra{\ss}burger}

\begin{document}

%\begin{frontmatter}
  \title{A fully labelled proof system for intuitionistic modal logics}
  \author{Sonia Marin \and %}
%  \todo{New university?}
%  \address{IT-Universitetet i K{\o}benhavn \\ Denmark }
%    \author{
      Marianela Morales %}
 \and
% \marianela{I think that now I have the same as Lutz, right?} No, the other order...
% \address{ LIX, \'Ecole Polytechnique  \& Inria Saclay\\ France}
%      \author{
        Lutz Stra{\ss}burger }
% \address{Inria Saclay \& LIX, \'Ecole Polytechnique \\ France}

  \maketitle

 \begin{abstract}

   In this paper we present a labelled sequent system for
   intuitionistic modal logics such that there is not only one, but
   two relation symbols for the labels: one for the accessibility
   relation associated with the Kripke semantics for normal modal logics and
   one for the preorder relation associated with the Kripke semantics
   for intuitionistic logic.
   %%
   This puts our system in close correspondence with the standard
   birelational Kripke semantics for intuitionistic modal logics. As a
   consequence it can encompass a wider range of intuitionistic modal
   logics than standard labelled systems.
   %%
   We also show an internal cut elimination proof for our system.

   %% . We show the possibility of extend labelled sequents
   %% with a preorder relation symbol in order to capture intuitionistic
   %% modal logic. For this, we obtain a proof system which is complete
   %% with respect to Hilbert system.  Also, we present the proof for
   %% completeness using Simpson system.
  \end{abstract}

  %% \begin{keyword}
  %% Intuitionistic modal logic, labelled sequents, cut elimination, proof theory.
  %% \end{keyword}
% \end{frontmatter}

  %\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{get bib-file right}

\section{Introduction}\label{sec:intro}

Since their introduction in the 1980s by Gabbay~\cite{gabbay:96},
\emph{labelled proof calculi} have been widely used by proof theorist
to give sound, complete and cut-free deductive systems to a broad range
of logics. Unlike so-called \emph{internal calculi}, like
hyper-sequents~\cite{avron:elc96} or nested
sequents~\cite{kashima:sl94,brunnler:aml09,poggiolesi:tmp09}, 2-sequents~\cite{masini:apal92}, or linear nested sequents~\cite{lellmann?},
labelled calculi have the advantage of being more uniform and being able to 
accommode more logics.

Standard labelled sequent calculi attach to every formla $\fm A$ a label
$\lb x$, witten as $\labels xA$, and additionally admit so called
\emph{relational atoms} of the form $\accs xy$ where $\rel$ is a
binary relation symbol. These calculi work best for logics with
standard Kripke semantics, as in this case the relation $\rel$ is used
to encode the accessibility relation in the Kripke models, and the
frame conditions corresponding to the desired logic can be directly
encoded as inference rules. Prominent examples are classical modal
logics and intuitionistic propositional logic, where, e.g., the frame
condition of transitivity ($\forall \lb {xyz}.\:\accs xy\AND \accs yz\IMP\accs xz$), can be straightforwardly translated into the
inference rule
\begin{equation}
  \label{eq:Rtrans}
  \vlinf{\rn{trans}}{}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs xz, \Left \SEQ \Right}
\end{equation}
where $\B$ stands for a set of relational atoms, and $\Gamma$ and
$\Delta$ for multi-sets of labelled formulas.

However,
in this paper we are concerned with \emph{intuitionistic modal
  logics}, whose Kripke semantics is based on \emph{birelational}
frames, i.e., they have two binary relations instead of one: one
relation $R$ that corresponds to the accessibility relation in
ordinary Kripke frames for modal logics, and a relation $\le$ that
corresponds to the future relation in ordinary Kripke frames for
intuitionistic logic. Consequently, standard labelled systems for
these logics have certain shortcomings:
\begin{enumerate}
\item The transitivity rule in~\eqref{eq:Rtrans} corresponds to both
  versions of the $\fourax$-axiom
  \begin{equation}
    \label{eq:4ax}
    \fm{\BOX A\IMP \BOX\BOX A}
    \qquand
    \fm{\DIA\DIA A\IMP\DIA A}
    \quadcm
  \end{equation}
  as they are equivalent in classical
  modal logic. However, in intuitionistic modal logic they are not
  equivalent, and even though the logic $\ISfour$, i.e., the
  intuitionistic version of the modal logic $\Sfour$, contains both
  axioms, they can be added independently to the logic $\IK$ (the
  intuitionistic version of $\K$). However, the proof theory of these
  new logics has not been studied before, as no existing labelled (or
  label-free) proof system can handle them, even though the
  corresponding frame conditions
  \begin{equation}
    \label{eq:4frame}
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{x'}.\:\futs x{x'}\AND\accs {x'}z)
    \quand
    \forall\lb{xyz}.\:\accs xy\AND \accs yz\IMP(\exists\lb{z'}.\:\futs z{z'}\AND\accs {x}{z'})
    \;,
  \end{equation}
  respectively, have already been studied
  in~\cite{plotkin:stirling:86}.
  \item The correspondence between the syntax and the semantics is not
  as clean as one would expect. As only the $R$-relation (and not the
  $\le$-relation) of the frame is visible in an ordinary labelled
  sequent, we only have that a sequent $\Gamma$ is provable if and
  only if is is valid in all \emph{graph-consistent}\footnote{This
    means that every layer in the model can be lifted to any future of
    any world in that layer. See~\cite{simpson:phd} and~\cite{mar:str:tableaux17} for a formal definition and discussion.} for  models, as
  already observed in by Simpson in his PhD thesis~\cite{simpson:phd}
  and considered as an inelegant solution (see also \cite{mar:str:tableaux17}).
\end{enumerate}
In order to address these two issues we propose here to extend the
notion of labelled sequent by also allowing relational atoms of the
form $\futs xy$. Consequently, we can easily translate the frame
conditions in~\eqref{eq:4frame} into inference rules:
\begin{equation}
  \label{eq:4rules}
  \vlinf{\rn{4^{\BOX}}}{\proviso{$\lb{x'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
     \B, \accs xy, \accs yz, \accs {x'}z, \futs x{x'},\Left \SEQ \Right}
  \qquand
  \vlinf{\rn{4^{\DIA}}}{\proviso{$\lb{z'}$ fresh}}{\B, \accs xy, \accs yz, \Left \SEQ \Right}{
    \B, \accs xy, \accs yz, \accs {x}{z'}, \futs z{z'},\Left \SEQ \Right}  
\end{equation}
This allows us to define cut-free deductive systems for a wide range
of logics that could not be treated before.  Furthermore, the relation
between syntax and semantics is as one would expect: A sequent is
provable in our system if and only if it is valid in all models.

Besides that, there is another pleasant observation to make about our
system. Ordinary labelled sequent systems for intuitionistic modal
logic are single-conclusion~\cite{simpson:phd}. The same is true for
the corresponding nested sequent
systems~\cite{str:fossacs13,marin:str:aiml}. It is possible to express
Maehara style multiple-conclusion systems in nested
sequents~\cite{kuz:str:maehara}, and therefore also in ordinary
labelled sequents. However, also in these systems there are rules
($\rlabrn\IMP$ and $\rlabrn\BOX$) that force a single-conclusion
premise, even though this is not the case in labelled
systems~\cite{negri:jpl2005} or nested sequents~\cite{fitting:83} for
intuitionistic logic. Our system eliminates this
discrepancy. Consequently, every rule in our system is invertible, as
we never delete information.

\medskip

This paper is organized as follows, In the next section (Section~\ref{sec:intmod}) we recall the standard syntax and semantics of intuitionistic modal logics. Then, in Section~\ref{sec:system} we present our system for the intuitionistic modal logic $\IK$. In Sections~\ref{sec:soundness} and~\ref{sec:completeness}, we show soundness and completeness of the system with cut. The cut elimination theorem, proved in Section~\ref{sec:cut-elim}, then entails soundness and completeness for the cut-free system. Finally, in Section~\ref{sec:ext} we discuss the possible extension to the system to capture other intuitionistic modal logics.

\todo{maybe paragraph on related work? cite Maffezioli, Naibo and Negri~\cite{maffezioli:etal:synthese13}.}

\section*{Introduction(old version, to be deleted)}
\todo{We should work on the introduction}

One possible-world semantics was established as a solid base to define modal logics, the idea of incorporating these notions into the proof theory of modal logics emerged. Fitch seems to have been the first one to formalise it, directly including symbols representing worlds into the language of his proofs in natural deduction \cite{Fitch}.

\emph{Labelled deduction} has been proposed by Gabbay~\cite{Gabbay} in the 80s as a unifying framework throughout proof theory in order to provide proof
systems for a wide range of logics. 
%
For modal logics it can take
the form of labelled natural deduction and labelled sequent systems as
used, for example, by Simpson~\cite{Simpson}, Vigan\`o~\cite{Vigano}, and
Negri~\cite{Negri}. 
%
These formalisms make explicit use not only of
labels, but also of relational atoms referring to the accessibility relation of a Kripke frame.
%
In this short note we propose a system that represents both the \emph{accessibility relation} (for modal
logics) and the \emph{preorder relation} (for intuitionistic
logic), using the full power of the bi-relational semantics for
intuitionistic modal logics,
and developing fully the idea that Maffezioli, Naibo and Negri employed in~\cite{Maffezioli}.

\todo{Simpson does not have internal cut-elimination}

\sonia{
  \begin{itemize}
  \item internal vs externl calculi
  \item extesion to axioms are truly intuitionistic (not just the
    classical versions as in Simpson style)
  \item connection with semantics compared to graph homomorphism
    restiction in Simpson and ``indexed nested systems''
  \end{itemize}
}

\lutz{things that should be said explicitely somewhere (maybe intro, maybe somewhere else):
  \begin{itemize}
  \item what we do:
    \begin{itemize}
    \item first systems (nested and labelled) for intuitionistic modal
      logic where ``single conclusion'', i.e., only one formula on the
      right.
    \item Then ``multiple conclusion'' (Meahara-style) nested sequent.
    \item Still rules that forced single formula on the right (white implication and white box)
    \item by allowing both relations of the model in the proof system
      we go to genuinely multiple conclusion: never ever a formula on
      the right has to be deleted during proof search.
    \end{itemize}
  \item why we do it: 
    \begin{itemize}
    \item getting things right: labelled systems for classical modal
      logics, or plain propositional intuiionistic logic have one
      binary relation, which is exactly the relation of the standard
      Kripke semantics. The standard Kripke semantics for
      intuitionistic modal logic is birelational, with a strong
      connectiion between the two binary relations. A clean labelled
      system for these logics should incorporate both relations.
    \item We get new logics.  In the standard nested sequent
      or labelled sequent system for intuitionistic modal logic we
      always have both conjuncts we we add Scott-Lemmon/Geach axioms;
      but here we can have them separtely.
    \end{itemize}
  \item how we do it:
    \begin{itemize}
    \item We start from the Maehara nested sequent system (which can
      be translated straightforwardly into a labelled system with one
      relation $R$.
    \item the rules for $\AND$ and $\OR$ are trivial and as expected
      they can be presented such that they are invertible on the left
      and the right, they do not concern the relations $R$ and $\le$. 
    \item So should be the rules for $\IMP$-left and $\BOX$-left and
      identity, but there is a subtlety; see below...
    \item the rules for $\DIA$ are also straightforward, as they
      concern only $R$ but not $\le$.
    \item The rules $\BOX$-right and $\IMP$-right are the ones that
      force single conclusion in the Maehara system. This means that
      here they create a future, using the relation $\le$.
    \item In order to make everything work together, this forces us to
      have reflexivity and transitivity of $\le$, explicely as rules
      in the system, and have rules for the F1 and F2 conditions
      making the relations $R$ and $\le$ go together.
    \item As a consequence we also need a rule realizing monotonicity
      for the formulas on the right.
    \item However, we designed the system such that monotonicity is
      admisible by encorporating it into the rules for identity,
      $\IMP$-left and $\BOX$-left.
    \end{itemize}
  \end{itemize}
}
%
%For sequent systems for intuitionistic logics there is always a choice
%to be made: make the system \emph{single conclusion} following
%Gentzen~\cite{Gentzen} or \emph{multiple conclusion} following
%Maehara~\cite{Maehara}. In our work we choose the multiple conclusion
%variant because of the closer correspondence to the semantics. In that
%respect, our system is closer to~\cite{kuz:str} than
%to~\cite{Simpson} and~\cite{mar:str}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
\section{Intuitionistic modal logics}\label{sec:intmod}
The language of {intuitionisitic modal logic} is the one of intuitionistic propositional logic with the modal operators $\BOX$ and $\DIA$, standing most generally for \emph{necessity} and \emph{possibility}.
%
Starting with a set $\mathcal{A}$ of atomic propositions, denoted by lower case letters $\fm a$, $\fm b$, $\fm c$,\ldots, modal formulas, denoted by capital letters $\fm A$, $\fm B$, $\fm C$, \ldots, are constructed from the grammar:
%
$$
\fm A \coloncolonequals
\fm a \mid \fm{A \AND A} \mid \fm{A \OR A} \mid \fm \BOT \mid \fm{A \IMP A} \mid \fm{\BOX A} \mid \fm{\DIA A}
$$
%

Obtaining the intuitionistic variant of $\mathsf{K}$ is more involved than the classical variant. 
%%
Lacking De Morgan duality, there are several variants of $\mathsf{k}$ that are classically but not intuitionistically equivalent. Five axioms have been considered as primitives in the literature. An intuitionistic variant of the modal logic $\K$ can then be obtained from ordinary intuitionistic propositional logic $\rn{IPL}$ by adding:
%

%The axiomatisation that is now generally accepted as intuitionistic modal logic $\IK$ was given by Plotkin and Stirling~\cite{Plotkin} and is equivalent to the one proposed by Fischer-Servi~\cite{Fischer}. It is obtained from intuitionistic propositional logic $\mathsf{IPL}$ by adding:
\begin{itemize}
	\item the \emph{necessitation rule}: if $\fm A$ is a theorem then $\fm{\BOX A}$ is also a theorem; and
	\item the following five variants of the \emph{distributivity axiom} $\rn{k}$:
	\begin{equation*}
	\label{eq:ik}%\hskip-2em
	\begin{array}[t]{r@{\;}l@{\qquad}r@{\;}l@{\qquad}r@{\;}l}
	\kax[1]\colon&\fm{\BOX(A\IMP B)\IMP(\BOX A\IMP\BOX B)}
	&
	\kax[3]\colon&\fm{\DIA(A\OR B)\IMP(\DIA A\OR\DIA B)}
	&
	\kax[5]\colon&\fm{\DIA\BOT\IMP\BOT}
	\\
	\kax[2]\colon&\fm{\BOX(A\IMP B)\IMP(\DIA A\IMP\DIA B)}
	&
	\kax[4]\colon&\fm{(\DIA A\IMP \BOX B)\IMP\BOX(A\IMP B)}\\%x[1ex]
	\end{array}
	\end{equation*}
\end{itemize}

The idea is that intuitionistic propositional logic does not allow the principle of \emph{Excluded Middle}, so the modalities $\BOX$ and $\DIA$ are not de Morgan duals any more, but one can choose to design the axiomatisation in order to relate them in different ways. The most basic intuitionistic modal system one can think of would be to consider only the $\BOX$ modality as regulated by the $\kax$ axiom (or as called here $\kax[1]$), which gives the system \hbox{$\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$}. However this would give strictly no information on the behaviour of the $\DIA$ modality.
It seems that Fitch \cite{fitch:pm48} was the first one to propose a way to treat $\DIA$ in an intuitionistic system by considering the system  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$, which is now sometimes called $\rn{CK}$ for \emph{constructive modal logic}. Wijekesera \cite{wijesekera:apal90} also considered the axiom $\kax[5]$, which states that $\DIA$ distributes over 0-ary disjunctions, but did not assume that it would always distribute over binary disjunctions; the system he proposed was therefore  $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$  $\rn{+}$ $\kax[5]$. The systems are well-designed for some applications for example to analyse some type systems \cite{benton:etal:jfp98} or to reason about states of a machine under partial information \cite{wijesekera:nerode:apal05}. However, they cannot satisfy the logicians as the addition of the \emph{Excluded Middle} principle to it does not yield classical modal logic $\K$, that is, it is not possible to retrieve the De Morgan duality of $\BOX$ and $\DIA$ in this case.

The axiomatisation that is now generally accepted as \emph{intuitionistic modal logic} denoted by $\IK$ was given by Plotkin and Stirling \cite{plotkin:stirling:86} and is equivalent to the one proposed by Fischer-Servi \cite{fischer-servi:84} and by Ewald \cite{ewald:jsl86} in the case of intuitionistic tense logic. It is taken to be    $\rn{IPL}$ $\rn{+}$ $\rn{nec}$ $\rn{+}$ $\kax[1]$ $\rn{+}$ $\kax[2]$ $\rn{+}$ $\kax[3]$ $\rn{+}$ $\kax[4]$  $\rn{+}$ $\kax[5]$.

The Kripke semantics for $\IK$ was first defined by Fischer-Servi~\cite{fischer-servi:84}.
%
It combines the Kripke semantics for intuitionistic propositional logic and the one for classical modal logic, using two distinct relations on the set of worlds.

\begin{definition}
	A \emph{bi-relational frame} $\F$ is a triple $\langle W, \rel, \le \rangle$ 
	%	of a non-empty set of worlds $W$ equipped with two binary relations $R$ and $\le$, where $R$ being the modal \emph{accessibility relation} and $\le$ a preorder (\emph{i.e.} a reflexive and transitive relation), satisfying the following conditions:
	of a set of worlds $W$ equipped with an {accessibility relation} $\rel$ and a preorder $\le$ (\emph{i.e.}\ a reflexive and transitive relation) satisfying:
	\begin{enumerate}
		\item[($\rn{F_1}$)] For all $\lb u, \lb v, \lb{v'} \in W$, if $\accs uv$ and $\futs v{v'}$, there exists $\lb{u'}$ s.t.~$\futs u{u'}$ and $\accs{u'}{v'}$.

		\begin{equation*}
		\xymatrix{
			\lb{u'} \ar@{.>}[r]^R  & \lb{v'} \\
			\lb u \ar@{.>}[u]^{\le} \ar@{->}[r]^R  & \lb v \ar@{->}[u]_{\le}
		}
		\end{equation*}

		\item[($\rn{F_2}$)] For all $\lb {u'}, \lb u, \lb v \in W$, if $\futs uv$, there exists $\lb {v'}$ s.t.~$\accs{u'}{v'}$ and $\futs v{v'}$.

		\begin{equation*}
		\xymatrix{
			\lb{u'} \ar@{.>}[r]^R & \lb{v'} \\
			\lb u \ar@{->}[u]^{\le} \ar@{->}[r]^R & \lb v \ar@{.>}[u]_{\le}
		}
		\end{equation*}
	\end{enumerate}
	%	
\end{definition}

\begin{definition}
	\label{model}
	A \emph{bi-relational model} $\M$ is a quadruple $\langle W, \rel,\le,V \rangle$ with $\langle W, \rel, \le \rangle$ a bi-relational frame and $V\colon W \to 2^\mathcal{A}$ a monotone valuation function, that is, a function mapping each world $\lb w$ to the subset of propositional atoms true at $\lb w$, additionally subject to:
	if $\lb w \le \lb{w'}$ then $V(\lb w)\subseteq V(\lb{w'})$.
\end{definition}

We write $\M, \lb w \Vdash \fm a$ if $\fm a \in V(\lb w)$, and inductively extend the $\Vdash$ relation to all formulas, following the rules for both intuitionistic and modal Kripke models:

\begin{equation}\label{eq:kripke}
	\begin{array}{rcl}
	%\parbox{30em}{%

		\M, \lb w \Vdash \fm{A \AND B} & \mbox{iff} & \M, \lb w \Vdash \fm A \mbox{ and } \M, \lb w \Vdash \fm B\\

		\M, \lb w \Vdash \fm{A \OR B} & \mbox{iff} & \M, \lb w \Vdash \fm A \mbox{ or } \M, \lb w \Vdash \fm B\\

		\M, \lb w \Vdash \fm{A \IMP B} & \mbox{iff} & \mbox{for all } \lb{w'} \mbox{ with } \futs w{w'}, \mbox{ if } \M, \lb{w'} \Vdash \fm A \mbox{ then } \M, \lb{w'} \Vdash \fm B\\

		\M, \lb w \Vdash \fm{\BOX A} & \mbox{iff} & \mbox{for all } \lb{w'} \mbox{ and } \lb u \mbox{ with } \futs w{w'} \mbox{ and } \accs {w'}u, \M, \lb u \Vdash \fm A \hfill \\ %$(\ast)$

		\M, \lb w \Vdash \fm{\DIA A} & \mbox{iff} & \mbox{there exists a } \lb u \mbox{ such that } \accs wu \mbox{ and } \M, \lb u \Vdash \fm A.

	\end{array}
\end{equation}
Observe that we never that $\M, \lb w \Vdash \fm\bot$. 
%
We write $\M, \lb w \not\Vdash \fm A$ if it is not the case that $\M, \lb w \Vdash \fm A$, but contrarily to the classical case, we do not have $\M, \lb w \Vdash \fm{\neg A}$ iff $\M, \lb w \not\Vdash \fm A$ (since $\fm{\neg A}$ is defined as $\fm{A \IMP \bot}$).

From the monotonicity of the valuation function $V$, we get a monotonicity property for the relation:
\begin{proposition}(Monotonicity)\label{prop:mon}
	For any formula $\fm A$ and for $\lb w, \lb{w'} \in W$, if $\futs w{w'}$ and $\M, \lb w \Vdash \fm A$, then $\M, \lb{w'} \Vdash \fm A$.
\end{proposition}

\begin{definition}
	A formula $\fm A$ is \emph{satisfied} in a model $\M = \langle W, \rel, \le, V \rangle$, if for all $\lb w \in W$ we have $\M, \lb w \Vdash \fm A$.
	%
	A formula $\fm A$ is \emph{valid} in a frame $\F = \langle W, \rel, \le \rangle$, if for all valuations $V$, the formula $\fm A$ is satisfied in $\langle W, R, \le, V \rangle$.
\end{definition}

%It then was investigated in detail in~\cite{Simpson}, in which strong arguments are given in favour of this axiomatic definition: 
%
%it allows for adapting to intuitionistic logic the standard embedding of modal logic into first-order logic, and also provides an extension of the standard Kripke semantics for classical modal logic to the intuitionistic case.

Similarly to the classical case, the correspondence between syntax and semantics for $\IK$ can be stated as follows.

\begin{theorem}[\cite{fischer-servi:84,plotkin:stirling:86}]\label{thm:plotkin}
	A formula $\fm A$ is a theorem of $\IK$ if and only if $\fm A$ is valid in every bi-relational frame.
\end{theorem}

\todo{Add something else about Fischer-Servi theorem?}\lutz{no, but we should add the Plotkin-Stirling theorem about the relation between axioms and frame conditions} \todo{add an intro to the next theorem}

%Plotkin and Stirling give proper correspondence results, but they must distinguish the two dual parts of each axiom and appeal to both relations of the frames,

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
\section{The system}\label{sec:system}
